---
title: "Ebola Analysis and Plotting"
Author: 'Emma McEvoy'
date: "2026-02-04"
output: html_document
---

```{r setup, include= FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(outbreaks)
library(incidence)
library(EpiEstim)
```
# Plotting with the incidence package
```{r load}
onset <- outbreaks::ebola_sim$linelist$date_of_onset

i <- incidence(onset)
plot(i)
```

# Change appearance
```{r apperance}
plot(i, color= 'lavender',alpha = 0.8, xlab='Date case reported')
```
# Add breaks on x-axis
```{r breaks}
plot(i, color= 'gold',alpha = 0.9, xlab='Date case reported', n_breaks= 10)
```
# Weekly incidence
```{r weeekly}
i_week <- incidence(onset, interval = 7) #weekly incidence
plot(i_week)
plot(i_week, labels_week = FALSE)
```
```{r fit}
my_fit <- fit(i)
plot(i, fit= my_fit)
```
The entire graph doesn't have exponential growth, it has exponential decay/decline about half way through the displayed period. Hence the fit is off. To solve this the split function should be used.

```{r splitfit}
split_fit <- fit(i, split = as.Date("2014-12-15"))
plot(i, fit= split_fit)
```

Using optim_fit_split, the optimal average fit for before and after can be found. (Selects best date)
```{r optimfit}
optim_split <- fit_optim_split(i)
optim_split
optim_split$split
```
```{r plotoptimfit}
optim_fit <- optim_split$fit
plot(i, fit = optim_fit)
```

# Plotting with epiEstim
```{r serial}
s_interval <- make_config(
  mean_si= 15.3,
  std_si= 9.3
)
```

```{r Re}
my_R <-estimate_R(i, method="parametric_", config = s_interval)
plot(my_R)
```
```{r plots}
plot(my_R, what= "R", legend = TRUE,
     options_R = list(col="red", xlab= "Date"))
```

# Using ggplot()
```{r tibble}
sex <- outbreaks::ebola_sim$linelist$gender
cases <- tibble(onset= onset, sex= sex)
cases
```

```{r histogram}
g <- cases %>% ggplot(aes(x= onset, fill = sex)) + geom_histogram()
g
```

# Add labels
```{r labels}
g <- g + labs(title = "Histogram of onset dates", x = "Dates")
g
```

# Change colours
```{r color}
g <- g + scale_fill_manual(values = c('pink','skyblue')) +
  theme_minimal(base_size = 16)
g
```

# Add legend
```{r legend}
g <- g + scale_fill_manual(values = c('pink','skyblue'),
  labels = c('Female', 'Male'))
g
```