---
title: "Pre-Seminar-Task"
date: "2026-01-28"
output: html_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(incidence)
library(EpiEstim)
```

# Change the Serial Interval Mean and STD and compare
```{r importData, echo = FALSE}
case_data <- read_csv("firstwave_London.csv")
case_data
```
```{r reverse, echo = FALSE}
case_data <- case_data %>% map_df(rev)
case_data
```
```{r incidence, echo = FALSE}
i <- case_data$date %>% incidence()
i
```
```{r fix, echo = FALSE}
i$counts <- case_data %>% select(newCasesBySpecimenDate) %>% as.matrix()
i$n <- case_data %>% select(newCasesBySpecimenDate) %>% sum()
```

## Before
```{r serial}
s_interval <- make_config(
  mean_si = 12.0,
  std_si = 5.2
)
```
```{r Re}
my_R <- estimate_R(i, method= "parametric_si", config = s_interval)
plot(my_R)
my_R$R
```
## Updating Mean and STD
```{r NEWserial}
new_s_interval <- make_config(
  mean_si = 6.0,
  std_si = 3.5 
)
```
```{r NEWRe}
new_my_R <- estimate_R(i, method= "parametric_si", config = new_s_interval)
plot(new_my_R)
new_my_R$R
```

By setting the mean to 6 and STD to 3.5, we can estimate the R number for the earlier date ranges as it now has enough data. Comparing the mean R value where the start date is 6 and end is 12, the new settings return a value which is around a third that of the original (9.2 vs 3.2). The same pattern applies to the STD and quantiles.

Changing the settings has no impact on the epidemic curve and the number of incidence.

When the mean and std is updated to 6 and 3.5, the SI distribution graph's frequency peaks around day 4 which is significantly earlier than with the previous mean and std as it peaked around day 10.

The estimated R graph for the previous settings show a significant spike around the 2nd week of March, where the mean R number increases to 25. In comparison the new settings had a less drastic spike where the mean increased to 6. This difference is significant as the greater a R number is from 1 suggests the epidemic is growing a greater rate.