---
title: "Fitting a model to endemic prevalence"
author: "Emma Davis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(deSolve)
```

### The SEIR model for rotavirus

The SEIR model separates a population into compartments:

* $S$: Susceptible
* $E$: Exposed (not yet infectious)
* $I$: Infectious
* $R$: Recovered/Removed

Let's consider rotavirus, which is mostly a childhood infection. We'll assume that recovered people have waning immunity (at rate $\theta$). The rate of change equations are given by:

$\Delta S= \theta R -\frac{\beta SI}{N}$

$\Delta E=\frac{\beta SI}{N} -\alpha E$

$\Delta I=\alpha E-\gamma I$

$\Delta R=\gamma I - \theta R$

$\Delta C = \alpha E$

where $C$ is a "dummy" variable representing cumulative cases becoming infectious, so we do not need to consider recoveries or deaths out of this compartment.

#### Setting up the model

**Task 1:** Write the rate of change equations into the model function.

Number of lines of additional code required: 6

```{r model}
SEIRmodel <- function(t, state, parameters){
  with(as.list(c(state, parameters)),{
    
    N <- S+E+I+R
    
    # Rate of change equations
    
    # Combine rates of change into a list
    
  })
}
```

**Task 2:** In Wales between 2009 and 2012, there were approximately 28 hospitalised cases of rotavirus per 100,000 population per year. Assuming there are 3 million (`=3e6`) people in Wales, convert this into: 

a. Annual hospitalisations (total cases, entire population)
b. Daily hospitalisations (total cases, entire population)

Assuming only 1 in 70 cases are hospitalised, what is the best estimate of

c. True annual incidence (total cases)
d. True daily incidence (total cases)

Number of lines of additional code required: 7-8

```{r calc_incidence}

# Your code here

```

Rotavirus infections recover within 3-7 days, so let's assume a mean of 5 days ($\gamma=1/5=0.2$), and it takes around 2 days to develop symptoms and become infectious ($\alpha=1/2=0.5$). We also know that prior infection confers 52% protection against being reinfected ($\phi = 0.52$) for around a year ($\theta = 1/365$).

**Task 3:** Create a vector with named entries for these fixed parameters: $\alpha$, $\gamma$, and $\theta$.

Number of lines of additional code required: 4

```{r params}

# Your code here

```

If the population size is 3 million, let's set up the initial number of exposed cases for our burn in period as equal to the daily incidence.

**Task 4:** Create a vector for the initial values, with $E=E_0=$ true daily incidence, $I=0$ and $R=0$. Remember to include $S$, $E$, $I$, $R$, and $C$! (Note: you could use any initial conditions here, except for $E=I=0$ and you will eventually reach the same equilibrium).

Number of lines of additional code required: 5-6

```{r initialvals}

# Your code here

```

To pick the time period, we're going to want to see roughly how long it takes for the value of $I$ to stabilize. Let's start by trying a period of 10 years (3650 days), but there's no need to output daily values, so we'll reduce this to every 10 days.

**Task 5:** Create a sequence of times that starts at 0, ends at 3650 days, taking time steps of length 1 day using `seq()`. Remember that if you don't know what the inputs are to a function, you can start typing and see what R suggests, or type `?seq` into the terminal for Help.

Number of lines of additional code required: 1

```{r period}

# Your code here

```

Now we need to run the model for an example value of $\beta$ and observe how long it takes to stabilise. 

**Task 6:** Try $\beta=0.3$ and look at the tail of the outputs -- is 10 years long enough for values to stabilise? Plot the value of $I$ over time.

Number of lines of additional code required: 8-12

```{r run_burnin}

# Your code here

```

**Task 7:** Write a distance function that 

* Takes as inputs: A value $\beta$, fixed parameters, initial conditions, and annual incidence calculated from data.
* Combines $\beta$ with the other parameters into a vector.
* Sets the burn in time sequence.
* Runs the model using these values.
* Calculates proportional daily incidence, $i$, then finds out what it is at equilibrium (Hint: use `tail(n = 1)` to get the last row of your output).
* Converts this to annual incidence for the entire population
* Outputs the distance between the equilibrium annual incidence and the data estimate.

Number of lines of additional code required: 9-12

```{r distance}
calc_dist <- function(beta, params, inits, dat){
  
  # Combine beta with other parameters
  
  
  # Set sequence of times
  
  
  # Run model using ode() and save as a tibble
  
  
  # Calculate annual incidence at equilibrium from C, remembering
  # a) If your time steps aren't 1 day long you'll need to rescale
  # b) Use tail(n = 1) to get the final value
  # c) to convert to a vector using as_vector() at the end if using %>%
  
  # Calculate the distance between model and data
  # Remember to use `abs()` to ensure distance is positive!
  
}
```

**Task 8:** Try a few values of $\beta$ in your function `calc_dist()` to get an idea of a sensible range and check your function works.

Number of lines of additional code required: 0

```{r testbeta}
calc_dist(0.3,
          fixed_params,
          initial_values,
          true_annual_incidence)
```

**Task 9:** Use the function `optimise()` to minimise this distance and find the optimal value of $\beta$ for some reasonable guess of upper and lower bounds. Hint: You will need to set your tolerance very small to get this to converge: try `tol = 1e-7` (i.e. $=0.0000001$) -- this describes how many decimal points to optimise $\beta$ to.

Number of lines of additional code required: 7-8

```{r fit}

# Your code here

```


**Task 9:** Run the model for this fitted value of $\beta$ with time steps of length 1, then 

* Use `mutate()` and `lag()` to create a daily incidence column. 
* What is the equilibrium annual incidence in the model (i.e. the tail value)? How does this compare to our estimate of true annual incidence from the data?

Number of lines of additional code required: 8-10

```{r validate}
# Combine beta from fit and fixed_params

# set times for 0-10 years, by = 1

# run model and save as a tibble

# use mutate() and lag(n = 365) to calculate annual incidence from C

# select final value of annual incidence

# compare to true annual incidence (from Task 2)

```

**Task 10:** Use `ggplot` to plot our simulated annual incidence `i` (from the model) and our estimated true annual incidence (from the data) on the same graph.

Number of lines of additional code required: 5-10

```{r plotfit}

# Your code here

```


#### Demo: Impact of Mass Vaccination Campaign

So now we can take the final outputs of the model as our starting values for $S$, $E$, $I$, and $R$ in our endemic scenario with constant prevalence, and then predict what will happen if we introduce a new intervention, e.g. a vaccine.

We first setup our conditions, similar to above, using our tail end values of our model that has reached equilibrium as our initial values of $S$, $E$, $I$ and $R$. If we're only interested in calculating new cases (or averted cases) then we can set $C = 0$.

```{r modelsetup}
initial_values <- out %>% tail(n = 1) %>% select(S, E, I, R) %>%
                  mutate(C = 0) %>%
                  as_vector()

parameters <- c(beta = fit_beta$minimum, fixed_params)

years <- 40
times <- seq(0, 365*years, by = 1)
```

Let's assume we do a **one-off** mass vaccination campaign, where we manage to vaccinate 60% of the population with a vaccine that is 90% effective (i.e. we effectively vaccinate $0.9\times 0.6=0.54, so 54% of the population) and we assume vaccine-induced immunity wanes at the same rate as infection-induced immunity, so we don't need to create a new model. This means we move 54% of susceptible people to the recovered class.

```{r vaccine}
psi <- 0.9
coverage <- 0.6
people_vaccinated <- (psi*coverage)*initial_values["S"]
initial_values["S"] <- initial_values["S"] - people_vaccinated
initial_values["R"] <- initial_values["R"] + people_vaccinated
```

Now we can run the model with this starting condition, assuming we conduct the mass vaccination at time $t=0$ and plot what happens to the annual incidence and cumulative incidence.

```{r run}
out_vac <- as_tibble(ode(y = initial_values,
                     times = times,
                     func = SEIRmodel,
                     parms = parameters))

out_vac <- out_vac %>% mutate(i = C - lag(C, 365))

out_vac %>% ggplot() +
  geom_line(aes(x = time/365, y = i)) +
  geom_line(aes(x = time/365, y = true_annual_incidence), 
            color = "darkorange",
            lty = 2) +
  xlab("time (years)") +
  ylab("annual incidence") +
  theme_minimal()
```

We see that incidence was suppressed for 15-20 years, but then rebounds back to equilibrium levels - this is because we haven't changed any aspect of the model or parameters, just completed a one-off shift of some susceptible people to the recovered compartment. However, we will still have reduced the cumulative incidence over the entire period.

```{r comparison}
total_cases <- out_vac %>% select(C) %>% # cumulative cases
               tail(n = 1) %>% as_vector()
poss_cases <- as_vector(model_annual_incidence)*years # if no vaccine

averted <- poss_cases - total_cases
averted
```

Therefore `r toString(round(averted,0))` cases were averted by the one-off mass vaccination program, even though in the long term the incidence returns to pre-vaccine equilibrium levels, this is equivalent to `r toString(round(averted/years,0))` cases per year across a roughly `r toString(years)` year period.

To have a more lasting effect, we would need to have regular vaccination, either at birth, at a fixed rate, or repeat mass vaccination. The first two (birth, fixed rate) would require a change of model equations. The last (repeat mass) would use the same model, but the model would be run for a fixed period, then vaccine applied by shifting people from $S$ to $R$, then run again for a fixed period, and so on.