 ---
title: "Week 2 Pair Coding: Using incidence and EpiEstim"
author: "Emma Davis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(incidence)
library(EpiEstim)
```

## The Data

First we import the data and assign it to an object, which we call `case_data`. I'm using `readRDS()` rather than `read_csv()` because the data in this instance is saved as an RData file (extension .RDS) containing only a vector of dates.

```{r import, echo=FALSE}
case_data <- readRDS('casedates_London.RDS')
```


## Using the `incidence` package

```{r incidence}
i <- case_data %>% incidence()
i
```
We can then look at our data using the `plot()` function.

```{r plot}
plot(i)
```

We can filter by date using two different methods:
1. Using square brackets and numerical index, e.g. `i[1:20]` will select the first twenty dates
2. Using specific date ranges, e.g. `subset(i, from = as.Date("2020-02-25"))` will select all dates after the 25th Feb 2020. We can also use both `to =` and `from =` to pick upper and lower bounds.

#### Task 1: Filtering data
- Create a new incidence object called `firstwave` by filtering to the first four months of the data (up to "2020-06-03").
- Create another incidence object called `secondwave` by choosing a suitable start and end date (hint: have a look at your incidence plot for the entire range of dates).

```{r Task1}
# write your code here
firstwave <- subset(i, to = as.Date("2020-06-03"))
secondwave <- subset(i, from = as.Date("2020-08-07"), to = as.Date("2021-04-03"))
plot(firstwave)
plot(secondwave)
```

## Estimating the growth rate and doubling time

We can estimate growth rates and doubling times using the `fit()` function (assuming an exponential growth/decline model), which also gives us the 95% confidence intervals.

#### Task 2: Fitting a model
- Estimate the growth rate and doubling time for the first half of your `firstwave` object by fitting an exponential growth model
- Estimate the growth rate and doubling time for the first half of your `secondwave` object by fitting an exponential growth model
- Use `plot()` to visually assess how good your model is for both cases

```{r Task2}
# write your code here
first_fit <- fit(firstwave, split = as.Date("2020-04-02"))
second_fit <-  fit(secondwave, split = as.Date("2020-12-29"))
plot(firstwave,fit =  first_fit)
plot(secondwave, fit= second_fit)
```

As well as filtering our dataset, we can also use the `split = as.Date("...")` argument to split the dataset into "before" and "after" and get estimates of the growth rate and doubling time for both periods. There's also the function `fit_optim_split()`, which will pick the optimal split date for the model.

#### Task 3: Fitting a split model
- Choose a date you think might be appropriate and use `fit()` and the `split =` argument to get estimates of the growth rate for before and after your chosen date.
- Use the `fit_optim_split()` function to fit the model and find the optimal date and the growth rates for before and after. How close is the optimal date to your guess? 

```{r Task3}
# write your code here
split_fit <- fit(i, split = as.Date("2021-05-03"))

optim_fit <- fit_optim_split(i)
optim_fit$split


```

## Using `EpiEstim` to estimate the effective R number

To estimate the R number we need to have some knowledge of the distribution of either the serial interval (onset to onset) or the generation interval (infection to infection). The `EpiEstim` package has a function `estimate_R()` that uses the serial interval and incidence data to estimate the real-time value of the effective R number.

First we need to configure the serial interval distribution settings for this calculation, using the `make_config()` function, by providing a mean and a standard deviation for the serial interval.

```{r serial}
s_interval <- make_config(
  mean_si = 12.0,
  std_si = 5.2
)
```

Next, we use this as an input in the function `estimate_R()`.

#### Example:

```{r Re}
my_R <- estimate_R(i,method="parametric_si",config=s_interval)
```

#### Task 4: Estimating R
- Estimate R for your `firstwave` and `secondwave` incidence objects
- Which date ranges have R greater than 1?
- Which date ranges have R less than 1?
- What do you think might have caused these changes in R?

```{r Task4}
# write your code here
firstwave_R <- estimate_R(firstwave,method="parametric_si",config=s_interval)
secondwave_R <- estimate_R(secondwave,method="parametric_si",config=s_interval)
plot(firstwave_R)
plot(secondwave_R)
```
The first wave starts with an R value of 7.5 before decreasing to around 0 in the middle of February 2020. The R number is greater than 1 between the end of February until mid April 2020 where it is less than 1. During the second wave the R value is greater than 1 until it briefly drops below 1 around November 2020.Around September 2020 and late March 2021 the R value reaches 1. The value is consistently below 1 around mid January 2021.

The rapid decrease in R around April 2020 during the first wave could be caused by the lock down implemented which would reduce the rate of new infections as people were only in contact with household members.
When the R decreased to less than 1 in November 2020 during the second wave, this also timed with a brief lock down period. However the sharp spike just after could align with the Christmas period where people were mixing. The decline in cases after could be due to the role out of a vaccine/ an immunity as well as better weather.