---
title: "Unit 6: Pre-Seminar Task"
subtitle: "Compare Stochastic and Deterministic Models"
author: "Emma Davis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(adaptivetau)
library(deSolve)
```

Recall that the SIR model rate of change equations are given by:

$\Delta S = -\frac{\beta S I}{N}$

$\Delta I = \frac{\beta S I}{N} - \gamma I$

$\Delta R = \gamma I$

### The Stochastic Model

#### 1. Define transitions

First we have to define the transitions. There are two events that can happen:

1. A susceptible individual can become infected: $-1$ from $S$, $+1$ to $I$.
2. An infected individual can recover: $-1$ from $I$, $+1$ to $R$.

We write this as a **list** of two vectors, where each vector represents one type of event.

```{r transitions}
transitions <- list(c(S = -1, I = +1), # infection
                    c(I = -1, R = +1)) # recovery
```

#### 2. Set parameters

Next we set the parameters and initial values of $S$, $I$, and $R$. Parameters need to be defined as a **list** and initial values as a **vector**.

```{r params_sto}
params_sto <- list(beta = 0.5,
                   gamma = 0.2)

initial_sto <- c(S = 9999,
                 I = 1,
                 R = 0)
```

#### 3. Define rates

Next, we define a function that calculates the rates at which our two events happen -- in the same order as we defined our transitions.

```{r rates}
SIRrates <- function(x, params, t){
  N <- x["S"]+x["I"]+x["R"] # total population size
  c(params$beta*x["S"]*x["I"]/N, # rate of infection
    params$gamma*x["I"]) # rate of recovery
  }
```

#### 4. Run the simulations

We use a `for` loop to iterate through `n` simulations, adding the outputs to `all_outs`, setting our random seed to allow reproducibility.

```{r runsims}
set.seed(20230301)
n <- 1000 # number of simulations to run

all_outs <- tibble(time = c(),  # create an empty tibble to store outputs
                   S = c(), 
                   I = c(),
                   R = c(), 
                   sim = c())

# for each simulation 1 to n
for(i in 1:n){
  # run the adaptive tau function, generating a single simulation output
  out <- ssa.adaptivetau(initial_sto,  # vector of initial conditions
                         transitions,  # list of possible transitions
                         SIRrates,     # function to calculate transition rates
                         params_sto,   # list of model parameters
                         tf = 100)     # time frame (days)
  
  out <- as_tibble(out) %>% mutate(sim = i) # label simulation number
  all_outs <- rbind(all_outs, out) # bind new simulation onto outputs
}

stochastic <- all_outs # relabel (optional, but useful for clarity)
```

#### 5. Plotting

We can plot all simulations by grouping by `sim` and using `geom_line`, for each of $S$, $I$, and $R$. Argument `alpha` represents transparency (0 = fully transparent, 1 = fully opaque).

```{r plotI}
all_outs %>% ggplot(aes(x = time, y = I, 
                        group = sim)) + geom_line(alpha = 0.05)
```

### The Deterministic Model

#### 1. Set parameters

We set our parameters as a **vector** using the `c()` function for combining values. These parameters want to be the same as the ones in our stochastic model.

```{r params}
params_det <- c(beta = 0.5, 
                gamma = 0.2) 
```

#### 2. Define your initial values

For ease of comparison we're going to look at $S$, $I$ and $R$ as numbers of people, rather than proportions, so $S+I+R=10000$.

```{r initial}
initial_det <- c(S = 9999,
                 I = 1,
                 R = 0)
```

#### 3. Write your model as a function

Now we take our rate of change equations and write them as a function, remembering to include $N$ as we're considering numbers of people.

```{r model}
SIRmodel <- function(t, state, parameters){
  with(as.list(c(state, parameters)),{
    
    N <- S+I+R # total population size
    
    # Rate of change equations
    dS <- -beta*S*I/N
    dI <- beta*S*I/N - gamma*I
    dR <- gamma*I
    
    # Combine rates of change into a list
    list(c(dS, dI, dR))
  })
}
```

#### 4. Pick your time intervals

Now we define how long we want to run the model for and how frequently (at what intervals) we want to save the values of $S$, $I$ and $R$. We ran the stochastic model for 100 days, so we'll do the same here.

```{r times}
times <- seq(0, 100, by = 1)
```

#### 5. Run the model

Now we have all the components we need, we can run the model using the `ode()` function from `deSolve`.

```{r run}
out_det <- ode(y = initial_det,
               times = times,
               func = SIRmodel,
               parms = params_det)

out_det <- as_tibble(out_det)
```

#### 6. Plotting

We can plot our deterministic outbreak using `ggplot()`. To use ggplot in the way it's intended, we first want to reformat our data using `gather()`. This will help us plot $S$, $I$ and $R$ on the same plot. Using `fct_relevel` (factor relevel) ensures R won't default to alphabetical order when processing the variable.

**Activity: Try commenting out lines 177-178 before running the next code chunk and observe the ordering of the legend and the allocation of colours.**

```{r gather}
deterministic <- out_det %>% gather(key = compartment, 
                                    value = value, 
                                    c("S", "I", "R")) %>%
  mutate(compartment = fct_relevel(compartment, 
                                   "S", "I", "R"))
deterministic 
```

Then we can 

```{r plot}
deterministic  %>%
            ggplot(aes(x = time, y = value, group = compartment)) +
            geom_line(aes(color = compartment,
                          linetype = compartment),
                          linewidth = 1) +
            scale_color_manual(values = c("darkgreen",
                                          "darkred",
                                          "darkgrey")) +
            theme_minimal(base_size = 12) +
            labs(x = "time (days)",
                 y = "proportion",
                 title = "The SIR model",
                 subtitle = "beta = 0.3, gamma = 0.2",
                 color = "",
                 linetype = "")
```

### Comparing the models

#### Combine results

We want to combine the results into one tibble. At the moment our two output tibbles are in different formats:

+ `stochastic` has columns: `time`, `S`, `I`, `R` and `sim`.
+ `deterministic` has columns: `time`, `compartment`, `value`.

So we need to use `gather()` to reformat our `stochastic` tibble, add a `sim` column to `deterministic` (this will just = 1001 because deterministic models only need to be run once and the first 1000 sims are stochastic) and add a column indicating whether a simulation is stochastic or deterministic.

```{r combining}
stochastic <- stochastic %>% gather(key = compartment, 
                                    value = value,
                                    c("S", "I", "R")) %>%
                             mutate(model = "stochastic")

deterministic <- deterministic %>% mutate(sim = 1001,
                                          model = "deterministic")

combined <- rbind(stochastic, deterministic) %>%
            mutate(compartment = fct_relevel(compartment,
                                             "S", "I", "R"))
```

Now we can plot both models and compare. We use `interaction()` when telling `ggplot` which factors to group by so that we can group by simulation number and compartment type. Using `alpha = model` and setting a manual scale for `alpha`, gives a solid line for the deterministic model and semi-transparent lines for the stochastic model runs.

```{r compare}
combined %>% ggplot(aes(x = time, y = value, 
                        group = interaction(sim, compartment))) +
             geom_line(aes(alpha = model, color = compartment)) +
             scale_alpha_manual(values = c(1, 0.01), 
                                guide = "none") +
             scale_color_manual(values = c("darkgreen",
                                          "darkred",
                                          "darkgrey")) +
             theme_minimal(base_size = 12) +
             labs(x = "time (days)",
                  y = "number of people",
                  title = "The SIR model",
                  subtitle = "beta = 0.3, gamma = 0.2",
                  color = "")
```

**Activity 2: What proportion of simulations don't result in an outbreak?**

Plot the stochastic simulations again and notice that the solid red line is below the horizontal line representing $I=20$.

```{r threshold}
stochastic %>% filter(compartment == "I") %>%
  ggplot(aes(x = time, y = value, group = sim)) +
  geom_line(alpha = 0.05, color = "darkred") +
  geom_hline(aes(yintercept = 20), lty = 2) +
  theme_minimal()
  
```

Using $I=20$ as a threshold for whether an outbreak will take-off, work out what proportion of simulations result in an outbreak:
+ Filter `stochastic` to 
  ++ only rows representing infectious values (`compartment == "I"`)
  ++ number of individuals (`value`) >20
+ Select column `sim`
+ Count the number of unique simulation numbers (using function `unique()` or otherwise)
+ Divide by the total number of simulations, `n`

```{r outbreaks}

```