---
title: 'Unit 1: Introduction to R'
author: "Emma McEvoy"
date: "2026-01-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


## Open dataset and save to variable ##

```{r open}
london_firstwave <- read_csv("firstwave_London-1.csv")
london_firstwave
```
## 1. Rename the columns ##

```{r rename}
london_firstwave <- london_firstwave %>% rename(newCases = newCasesBySpecimenDate, cumCases = cumCasesBySpecimenDate, firstEpisode = newFirstEpisodesBySpecimenDate, cumFirstEpisode = cumFirstEpisodesBySpecimenDate,newReinfections = newReinfectionsBySpecimenDate,cumReinfections= cumReinfectionsBySpecimenDate)
london_firstwave
```
## 2. Check that the columns containing "FirstEpisode" data are equal to your newCases and cumCases columns ##
```{r equal}
#new <- london_firstwave %>%filter(newCases==firstEpisode) 
new <- london_firstwave %>%filter(cumCases==cumFirstEpisode)
new
```
Filtering by firstEpisode returns 61 rows which is the original table length so none are missing/ different.

## 3. Check that columns containing "Reinfections" data are empty##
```{r empty}
london_firstwave %>% select(newReinfections, cumReinfections)
```
## 4. Remove columns from data
```{r remove}
london_firstwave <- london_firstwave %>% select(date, newCases, cumCases)
london_firstwave
```
## 5. Sort data
```{r sortNewCases}
sortByNewCases <- london_firstwave %>% arrange(newCases)
sortByNewCases
```
```{r sortDate}
sortByDate <- london_firstwave %>% arrange(newCases)
sortByDate
```
## 6. Filter to only days where "newCases" is greater than 50##
```{r filter}
over50 <- london_firstwave %>% filter(newCases>50)
over50
```