---
title: "Unit 2: Analysing Epidemiological Data"
output: html_document
date: 
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(incidence)
library(EpiEstim)
```

# Import the data
```{r importData}
case_data <- read_csv("firstwave_London.csv")
case_data
```

# Reverse data
```{r reverse}
case_data <- case_data %>% map_df(rev)
case_data
```

# Using Incidence function
```{r incidence}
i <- case_data$date %>% incidence()
i
```
As the number of cases on a given date are stored under the variable 'newCasesBySpecimenDate' this function incorrectly assumes there's only one case per day. 

# Replacing the counts with newCaseBySpecimenDate column data
```{r fix}
i$counts <- case_data %>% select(newCasesBySpecimenDate) %>% as.matrix()
i$n <- case_data %>% select(newCasesBySpecimenDate) %>% sum()
plot(i)
```

#Filtering dates
```{r dates1}
i_early <- i[1:20]
i_early
plot(i_early)
```
This filters to show the 1st 20 dates using square brackets. The alternative way to do this is by using specific date ranges and the subset function.
```{r dates2}
i_mid <- subset(i, from = as.Date("2020-02-25"), to = as.Date("2020-03-10"))
i_mid
plot(i_mid)
```

# Estimate the growth rate

```{r fit}
my_fit <- fit(i)
my_fit
```

This also gives us our 95% confidence interval. We can also split the data into before and after a specified date to get estimates of the different growth rates.
```{r split}
split_fit <- fit(i, split = as.Date("2020-03-01"))
split_fit
```

# Using EpiEstim to predict R number
```{r serial}
s_interval <- make_config(
  mean_si = 12.0,
  std_si = 5.2
)
```

This provides a mean and standard deviation for the serial interval. We use this as the input function for estimate_R()

```{r Re}
my_R <- estimate_R(i, method= "parametric_si", config = s_interval)
plot(my_R)
my_R$R
```