---
title: "Fitting a deterministic SIR model to toy data"
author: "Emma Davis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(deSolve)
library(adaptivetau)
```

#### The Data

The prevalence data in `SimulatedOutbreak.RDS` has been generated using an $SIR$ model with certain values of the transmission rate $\beta$ and recovery rate $\gamma$, simulating an outbreak of a novel disease in a population of size $N=10000$.

```{r popsize}
N <- 10000
```

$\Delta S = -\frac{\beta I S}{N}$
$\Delta I = \frac{\beta S I}{N} - \gamma I$
$\Delta R = \gamma I$

We recall that the basic reproduction number, $R_0$, for the $SIR$ model is given by: $$R_0=\frac{\beta}{\gamma}$$

**Task 1:** Load the data using `readRDS(file = "filename.RDS")` and

a. Plot the number of active cases ($I$) over time

Number of lines of code required: 1

```{r Task1a}
# outbreak_data <- ??
```

b. What is the peak prevalence? When does this occur?

Number of lines of code required: 2

```{r Task1b}
# your code here
```

c. How long does it take until the end of the epidemic (0 cases)?

Number of lines of code required: 1-2

```{r Task1c}
# your code here
```

Next, we want to fit the $SIR$ model to the data to estimate values of $\beta$ and $\gamma$.

**Task 2**: Write a function for calculating the $SIR$ model rate of change equations (without births or deaths).

Number of lines of code required: 5

```{r SIRmodel}
SIR <- function(t, state, parameters){
  with(as.list(c(state, parameters)),{
    
    # Calculate N
    
    
    # Rate of change equations
    
    
    # Combine rates of change into a list
    
  })
}
```


**Task 3**: Write a function for calculating the sum of the squared errors for your model function for inputs `params` and `data`.

Hint: the output of your function is the item on the last line (so long as it isn't assigned to an object name).

Number of lines of code required: 6-8

```{r sse_SIR}
sse_SIR <- function(params, data){
  
  cases <- data %>% select(cases) %>% as_vector() %>% as.integer()
  
  # Pick model timesteps, t (to match the data)
  
  # Create a vector of named parameters (using input params)
  
  # Set your initial conditions, using the first element of cases, cases[1]
  
  # Use ode() to solve your model with these parameters, timesteps
  # and initial conditions, then save it as a tibble.
  
  # Calculate the sum of the squared errors
  
}
```

**Task 4:** Using `optim()`, fit your model to the data to find parameter estimates for $\beta$ and $\gamma$

Number of lines of code required: 3-6

```{r fit}
# Your code here
```

**Task 5:** Repeat Task 4 with your fitted outputs as your new starting values to double check convergence.

Number of lines of code required: 3-6

```{r checkfit}
# Your code here
```

**Task 6:** Run your model with the fitted parameter values.

Number of lines of code required: 6-9

```{r run}

# initial conditions (vector with named entries, from data)

# parameters (vector with named entries, from 2nd fit)

# output times (want these to match the data observations)

# Use ode() to solve your model with these parameters, timesteps
# and initial conditions, then save it as a tibble.
```

**Task 7:** Plot your model and the data on the same figure. For the model, calculate the peak prevalence and when it occurs -- compare this to your results from Task 1b.

Number of lines of code required: 2-4

```{r plot}
# Your code here
```